Supplementary Information for: Crosslinguistic Word Orders Enable an Efficient Tradeoff between Memory and Surprisal

### 2020-02-15 writeup/control-ngrams.tex
It plot

% ../../writeup/tables/quantiles_REAL_NGRAMS_0.tex

\begin{center}
\begin{longtable}{cccccccccccccccccc}
\input{tables/quantiles_REAL_NGRAMS_0.tex}
\input{tables/quantiles_REAL_NGRAMS_1.tex}
\input{tables/quantiles_REAL_NGRAMS_2.tex}
\end{longtable}
	\captionof{figure}{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language? Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds. At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less. Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
\end{center}

%bounds.append(["alpha", float, 0.95, 1.0]) # + [x/20.0 for x in range(15, 21)])
%bounds.append(["gamma", int, 1, 2, 3, 4, 5, 8, 10, 15, 20, 25, 30]) # , 200, 300
%bounds.append(["delta", int, 0.1, 0.2, 0.5, 1.0 , 2.0, 3.0, 4.0, 5.0, 8.0, 10.0]) #, 1024]) #, 1024]) # 64, 128,
%bounds.append(["cutoff", int, 2,3,4,5,6,7,8,9,10]) #,7,8,9,10]) #, 1024]) #, 1024]) # 64, 128,
%

>>>>

% ../../writeup/tables/quantiles_REAL_NGRAMS_0.tex
%
%\begin{center}
%\begin{longtable}{cccccccccccccccccc}
%\input{tables/quantiles_REAL_NGRAMS_0.tex}
%\input{tables/quantiles_REAL_NGRAMS_1.tex}
%\input{tables/quantiles_REAL_NGRAMS_2.tex}
%\end{longtable}
%	\captionof{figure}{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language? Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds. At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less. Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
%\end{center}
%
%bounds.append(["alpha", float, 0.95, 1.0]) # + [x/20.0 for x in range(15, 21)])
%bounds.append(["gamma", int, 1, 2, 3, 4, 5, 8, 10, 15, 20, 25, 30]) # , 200, 300
%bounds.append(["delta", int, 0.1, 0.2, 0.5, 1.0 , 2.0, 3.0, 4.0, 5.0, 8.0, 10.0]) #, 1024]) #, 1024]) # 64, 128,
%bounds.append(["cutoff", int, 2,3,4,5,6,7,8,9,10]) #,7,8,9,10]) #, 1024]) #, 1024]) # 64, 128,
%

### 2020-02-15 writeup/si.tex
It plot

\subsection{Median Surprisal per Memory Budget}

\begin{center}
\begin{longtable}{ccccccccccccccclll}
\input{tables/medians_REAL_0.tex}
\end{longtable}
	\captionof{figure}{Medians: For each memory budget, we provide the median surprisal for real and random languages. Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median, dotted lines indicate empirical quantiles ($10\%, 20\%, \dots, 80\%, 90\%$). Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
\end{center}

\begin{center}
\begin{longtable}{ccccccccccccccclll}
\input{tables/medians_REAL_1.tex}
\end{longtable}
	\captionof{figure}{Medians (cont.)}
\end{center}

\begin{center}
\begin{longtable}{ccccccccccccccclll}
\input{tables/medians_REAL_2.tex}
\end{longtable}
	\captionof{figure}{Medians (cont.)}
\end{center}

\begin{center}
\begin{longtable}{ccccccccccccccclll}
\input{tables/medians_REAL_3.tex}
\end{longtable}
	\captionof{figure}{Medians (cont.)}
\end{center}

\subsection{Surprisal at Maximum Memory}

\begin{center}
\begin{longtable}{ccccccccccccccclll}
\input{tables/slice-hists_REAL_0.tex}
\input{tables/slice-hists_REAL_1.tex}
\input{tables/slice-hists_REAL_2.tex}
\input{tables/slice-hists_REAL_3.tex}
\end{longtable}
	\captionof{figure}{Histograms: Surprisal, at maximum memory.}\label{tab:slice-hists-real}
\end{center}

@@

\subsection{Samples Drawn (Experiment 3)}

\begin{center}
\begin{tabular}{l|ll||l|llllllllllllll}
	Language & Base. & MLE & Language & Base. & MLE \\ \hline
\input{tables/samplesNumber_ground.tex}
\end{tabular}
	\captionof{figure}{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
\end{center}

\subsection{Medians (Experiment 3)}

\begin{center}
\begin{longtable}{ccccccccccccccclll}
\input{tables/medians_0.tex}
\input{tables/medians_1.tex}
\input{tables/medians_2.tex}
\input{tables/medians_3.tex}
\end{longtable}
	\captionof{figure}{Experiment 3. Medians: For each memory budget, we provide the median surprisal for real and random languages. Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median. Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
\end{center}

\begin{center}
\begin{longtable}{ccccccccccccccccll}
\input{tables/medianDiff_0.tex}
\input{tables/medianDiff_1.tex}
\input{tables/medianDiff_2.tex}
\input{tables/medianDiff_3.tex}
\end{longtable}
	\captionof{figure}{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red). Lower values indicate lower surprisal compared to baselines. Solid lines indicate sample means. Dashed lines indicate 95 $\%$ confidence intervals.}\label{tab:median_diffs}
\end{center}

\begin{center}
\begin{longtable}{cccccccccccccccccc}
\input{tables/quantiles_noAssumption_0.tex}
\input{tables/quantiles_noAssumption_1.tex}
\input{tables/quantiles_noAssumption_2.tex}
\end{longtable}
	\captionof{figure}{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language? Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds. At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less. Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
\end{center}

@@

\section{Optimization}
We cannot optimize for AUC, because we cannot construct an unbiased gradient estimator
$AUC \propto \sum_{t} (\sum_{s \leq t} I_s) tI_t = \sum_{s \leq t}  tI_t I_s$

>>>>

%\subsection{Median Surprisal per Memory Budget}
%
%\begin{center}
%\begin{longtable}{ccccccccccccccclll}
%\input{tables/medians_REAL_0.tex}
%\end{longtable}
%	\captionof{figure}{Medians: For each memory budget, we provide the median surprisal for real and random languages. Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median, dotted lines indicate empirical quantiles ($10\%, 20\%, \dots, 80\%, 90\%$). Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
%\end{center}
%
%\begin{center}
%\begin{longtable}{ccccccccccccccclll}
%\input{tables/medians_REAL_1.tex}
%\end{longtable}
%	\captionof{figure}{Medians (cont.)}
%\end{center}
%
%\begin{center}
%\begin{longtable}{ccccccccccccccclll}
%\input{tables/medians_REAL_2.tex}
%\end{longtable}
%	\captionof{figure}{Medians (cont.)}
%\end{center}
%
%\begin{center}
%\begin{longtable}{ccccccccccccccclll}
%\input{tables/medians_REAL_3.tex}
%\end{longtable}
%	\captionof{figure}{Medians (cont.)}
%\end{center}
%

%
%\subsection{Surprisal at Maximum Memory}
%
%
%\begin{center}
%\begin{longtable}{ccccccccccccccclll}
%\input{tables/slice-hists_REAL_0.tex}
%\input{tables/slice-hists_REAL_1.tex}
%\input{tables/slice-hists_REAL_2.tex}
%\input{tables/slice-hists_REAL_3.tex}
%\end{longtable}
%	\captionof{figure}{Histograms: Surprisal, at maximum memory.}\label{tab:slice-hists-real}
%\end{center}
%

@@

%
%
%\subsection{Samples Drawn (Experiment 3)}
%
%
%
%\begin{center}
%\begin{tabular}{l|ll||l|llllllllllllll}
%	Language & Base. & MLE & Language & Base. & MLE \\ \hline
%\input{tables/samplesNumber_ground.tex}
%\end{tabular}
%	\captionof{figure}{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
%\end{center}
%
%
%
%\subsection{Medians (Experiment 3)}
%
%\begin{center}
%\begin{longtable}{ccccccccccccccclll}
%\input{tables/medians_0.tex}
%\input{tables/medians_1.tex}
%\input{tables/medians_2.tex}
%\input{tables/medians_3.tex}
%\end{longtable}
%	\captionof{figure}{Experiment 3. Medians: For each memory budget, we provide the median surprisal for real and random languages. Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median. Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
%\end{center}
%
%
%
%
%
%\begin{center}
%\begin{longtable}{ccccccccccccccccll}
%\input{tables/medianDiff_0.tex}
%\input{tables/medianDiff_1.tex}
%\input{tables/medianDiff_2.tex}
%\input{tables/medianDiff_3.tex}
%\end{longtable}
%	\captionof{figure}{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red). Lower values indicate lower surprisal compared to baselines. Solid lines indicate sample means. Dashed lines indicate 95 $\%$ confidence intervals.}\label{tab:median_diffs}
%\end{center}
%
%
%
%
%
%
%
%\begin{center}
%\begin{longtable}{cccccccccccccccccc}
%\input{tables/quantiles_noAssumption_0.tex}
%\input{tables/quantiles_noAssumption_1.tex}
%\input{tables/quantiles_noAssumption_2.tex}
%\end{longtable}
%	\captionof{figure}{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language? Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds. At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less. Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
%\end{center}
%
%

@@

\section{Optimization}
We cannot optimize for AUC using the method of (CITE), because we cannot construct an unbiased gradient estimator
$AUC \propto \sum_{t} (\sum_{s \leq t} I_s) tI_t = \sum_{s \leq t}  tI_t I_s = I_1^2 + 2 I_1 I_2 + 2 I_2^2 + \dots$

As a surrogate, we propose to maximize $I_1$.
This corresponds

It provides an accurate approximation to the AUC if $I_s$ is small for $s > 1$, which holds for the n-gram based estimator.

If $I_s < \epsilon$ for $s > 1$, then 
\begin{equation}
	I_1^2 \leq \sum_{1\leq s \leq t \leq T}  tI_t I_s \leq I_1^2 + I_1 \epsilon T^2 + \epsilon^2 T^2
\end{equation}
That means, $I_t^2$ is an accurate approximation of the AUC when $\epsilon T^2$ is small.
